---
title: "Analysis of Vitamin C and tooth growth in Hamsters"
author: "Danny Andersen"
date: "22 October 2015"
output: pdf_document
---

```{r libraries,warning=FALSE, message=FALSE ,echo = FALSE}
library(ggplot2)
library(dplyr)
library(lattice)
```

##Synopsis

##Data Preparation

```{r dataprep}
data("ToothGrowth")
ham_id = { id<- NULL;for (i in 1:6) { for (i in 1:10) { id <- c(id,i)}}; id}
tgi <- cbind(ham_id, ToothGrowth)
```

```{r lattice_plot}

xyplot(len~dose | factor(ham_id), data=tgi, groups=supp, type="b", pch=c(19,0), 
       grid=T, col=c("blue","red"), 
       key=list(text=list(c("Orange Juice", "Vitamin C")),
                points=list(pch=c(19,0)),col=c("blue","red")),
       main="Hamster tooth length by Ascorbic Acid dosage and Supplent type for 10 hamsters", xlab="Dosage (mg)", ylab="Tooth length (mm)")
```

```{r overlay_plot }
tgif <- dplyr::filter(tgi,supp=="VC") %>% mutate(hamster = factor(ham_id))
g <- ggplot(tgif, aes(x=dose, y=len, group = hamster))
g <- g + labs(x="Dosage (mg)", y="Tooth length (mm)", title="Hamster tooth length by Ascorbic Acid dosage by Vitamin C supplement for 10 Hamsters")
g + geom_line(size =1, aes(colour=hamster))+ geom_point(pch = 19, fill = "salmon", alpha = .5)
```

##Does Ascorbic acid increase tooth growth in hamsters?

This is just one of the many questions about hamsters that you wanted to know but were afraid to ask.

The null hypothesis is that increasing the dosage of vitamin C doesn't increase tooth size, the alternate hypothesis is that is does.

From the graphs generated during the data analysis, it would appear that the largest tooth growth is when the dosage is administered through an orange juice supplement. As there is a low sample number (10), we will use the t-test interval to determined the probability of this alternate hypothesis.

```{r oj_ttest}
tg_05_oj <- filter(tgi, supp=="OJ" & dose==0.5)$len
tg_10_oj <- filter(tgi, supp=="OJ" & dose==1)$len 
tg_20_oj <- filter(tgi, supp=="OJ" & dose==2)$len 
t1<-t.test(tg_10_oj,tg_05_oj, paired=T)
t2<-t.test(tg_20_oj,tg_10_oj, paired=T)
```

The lattice graphs show a marked increase in tooth growth between 0.5mg and 1.0mg, so we will consider the difference of these two results first.

The 95% confidence interval of the difference in means of the alternate hypothesis is `r t1$conf.int`, with a t-statistic of `r t1$statistic` and a p-value of `r t1$p.value` of the null hypothesis. This shows that there is a high probability that a supplement of orange juice does increase tooth growth in hamsters. If we look at the effect of increasing the dose to 2mg from 1mg, the effect is in- conclusive, with the 95% confidence interval including 0: `r t2$conf.int`. 

However, there may be factors other than the ascorbic acid in the orange juice that is having this effect. So let's turn our attention to the pure vitamin C supplement.

```{r oj_ttest}
tg_05_vc <- filter(tgi, supp=="VC" & dose==0.5)$len
tg_10_vc <- filter(tgi, supp=="VC" & dose==1)$len 
tg_20_vc <- filter(tgi, supp=="VC" & dose==2)$len 
t3<-t.test(tg_10_vc,tg_05_vc, paired=T)
t4<-t.test(tg_20_vc,tg_10_vc, paired=T)
```

The correlation between tooth growth and vitamin C dosage is even greater when the dosage is through a vitamin C supplement. Looking at the difference between 0.5mg and 1.0mg, the 95% confidence interval of the difference in means of the alternate hypothesis is `r t3$conf.int`, with a t-statistic of `r t3$statistic` and a p-value of `r t3$p.value` of the null hypothesis. Increasing the dosage further from 1.0mg to 2.0mg confirms the alternate hypothesis, with a difference in mean confidence interval of `r t4$conf.int`

##Conclusion

There is strong statistical support of the hypothesis that ascorbic acid increases tooth growth in hamsters.

In addition, the lack of increase from 1.0mg to 2.0mg in tooth growth when delivered using orange juice but a strong increase when a pure vitamin C supplement was suggested that other factors came into play by increasing the dosage of vitamin C, e.g. increased tooth decay.

However, without a control group, i.e. where no vitamin C supplement is given, these findings cannot be conclusive, as simply a more nutrious diet may increase tooth growth, rather then a particular vitamin. We are also making the assumption that all hamsters have been treated equally, e.g. fed on the same diet and kept in the same conditions. 
